---
title: "Respiratory Disease Outbreaks in Toronto Pre, During, and Post Covid-19 Pandemic "
subtitle: How Coronavirus and Other Respiratory Diseases have responded to the covid-19 pandemic and Toronto pandemic policies
author: 
  - Evelyn Hughes
thanks: "Code and data are available at: [https://github.com/EveHughes/Donaldson-Paper)."
date: 05/12/2025
date-format: long

abstract: "Through the analysis of Outbreaks in Toronto Healthcare Institutions, this paper seeks to identify the trends in respiratory infections in Toronto during the implementation and later lifting of covid-19 preventatory policies, including general trends in number of outbreaks, and makeup of respiratory infections. Through data visualization, this analysis finds that while reported coronavirus outbreaks have lowered since pandemic times, “secondary” respiratory diseases such as Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus that were at lower levels during the pandemic have begun to rise again. At the same time, covid becomes a lesser percentage of overall respiratory diseases. Given different respiratory diseases require different treatment, these findings are relevant for informing healthcare facilities to place a lesser emphasis on covid-19, etc."

format: pdf
number-sections: true
bibliography: references.bib
---

```{python}
#| label: fig-polar-1
#| fig-cap: "A line plot on a polar axis"

import numpy as np
import matplotlib.pyplot as plt

r = np.arange(0, 2, 0.01)
theta = 2 * np.pi * r
fig, ax = plt.subplots(
  subplot_kw = {'projection': 'polar'} 
)
ax.plot(theta, r)
ax.set_rticks([0.5, 1, 1.5, 2])
ax.grid(True)
plt.show()
```


# Introduction

On January 25, 2020, the first case of COVID-19 was reported in Canada, in a healthcare facility in Toronto (Cheese 2025). Around two months later on March 17, Doug Ford declared a state of emergency for Ontario, and the closure of facilities such as schools. Additionally, mask mandates, social distancing guidelines, and vaccine cards were all put into place to mitigate the spread of the disease. Later around 2022, these restrictions began to lift. Schools went back in person, etc, etc, etc. (Compartin 2022). As the restrictions lift, it becomes important to characterize the effect they hold upon respiratory diseases both as a whole, and specific diseases.

Previously, in 2021, research was conducted to examine the effects of covid protocol on other respiratory diseases, finding evidence for a decrease in disease during covid protocol (Groves et al 2021). However, there is a lack of literature examining the effects of the lifting of protocol on these same diseases, and examining the general trends. This research aims to fill that gap, and identify how diseases have reacted to the lifting of covid protocol in relation to both their pre covid (2019), and during covid (2020-2022 ish) levels. Doing such can inform policy that encourages disease-specific preventative measures, and promotion of vaccines to prevent increased outbreaks.

This analysis first examines the trends in the overall number of outbreaks and respiratory outbreaks each year to better understand how the number of outbreaks has been varying over the years and identify general trends. Then, it looks at coronavirus, Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus. Finally, it examines the trends in percentage of respiratory outbreaks with covid as an agent. Through the data visualization, it was found that while coronavirus outbreaks have been falling since their covid levels, other respiratory diseases have been seeing a subsequent increase in cases post pandemic. Additionally, the percentage of respiratory outbreaks with covid as an agent has been decreasing since the pandemic. So, while an emphasis on covid specific protocols is important, preventative measures for other respiratory diseases, such as RSV vaccines, that were overshadowed during the pandemic should be further encouraged (Public Health Agency of Canada 2024).


In the sections that follow are a description of the data, visualization of the data and discussion of the results.


# Data {#sec-data}

## Raw Data

Outbreaks in Toronto Healthcare Institutions data is provided by Toronto Public Health., and is obtained from the OpenData Toronto Portal. The dataset is updated weekly on Thursdays with data as of the Wednesday at 2pm.(https://open.toronto.ca/dataset/outbreaks-in-toronto-healthcare-institutions/ ). 

The dataset consists of data from years 2016-2025 inclusive, with each containing instances of outbreaks in toronto healthcare institutions. 

“Under the Ontario Health Protection and Promotion Act (HPPA), healthcare institutions (hospitals, long-term care homes and retirement homes) are required to monitor staff and patients/residents for signs and symptoms of gastroenteric (e.g., nausea, vomiting, diarrhea, fever) and respiratory (e.g., cough, runny nose, sore throat, fever) infections. Healthcare institutions must also actively look for, detect and report suspected and/or confirmed outbreaks to their local public health unit”

For every record there is an id number, institution name, institution address, outbreak setting (type of facility), type of outbreak, causative agent-1, causative agent-2, date outbreak began, date outbreak ended, and whether the outbreak is currently active. 

The type of outbreak is listed as Respiratory, Enteric, or Other.
By the Ministry of Health, a respiratory outbreak is constituted by “Two cases of acute respiratory infections (ARI) within 48 hours with any common epidemiological link (e.g., unit, floor), at least one of which must be laboratory-confirmed; OR Three cases of ARI (laboratory confirmation not necessary) occurring within 48 hours with any common epidemiological link (e.g., unit, floor). https://www.toronto.ca/wp-content/uploads/2023/06/9942-Outbreak-Preparedness-ToolkitRespiratoryTableTopFacilitators.pdf

Causative Agent-1 is listed as either as the specific causative agent (eg Covid-19), or is listed as "Unable to Identify". 
Causative Agent-2 is listed as either the specific secondary causative agent, or is listed as "None".

Below is a preview of this information:

```{python}
#| label: fig-polar
#| fig-cap: "A line plot on a polar axis"

import numpy as np
import matplotlib.pyplot as plt

r = np.arange(0, 2, 0.01)
theta = 2 * np.pi * r
fig, ax = plt.subplots(
  subplot_kw = {'projection': 'polar'} 
)
ax.plot(theta, r)
ax.set_rticks([0.5, 1, 1.5, 2])
ax.grid(True)
plt.show()
```


For this analysis, the variables of interest are the yearly number of outbreaks for Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus over the years of 2019 - 2024 inclusive.

## Analysis Data
	
For each year, the number of cases were aggregated, and were aggregated within each respective category. The diseases of interest were: Coronavirus Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus. 
Instances of coronavirus and covid-19 were grouped together under the title of “coronavirus”. 
Influenza A & B were additionally grouped together under the title “Influenza”. 

Below is a sample of the cleaned data:


```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

And the summary statistics for the cleaned data:

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

# Results

## Total Outbreak, Coronavirus Outbreak, and Respiratory Outbreak 
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

## Influenza Outbreak & Other 

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

## Other outbreaks

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults-2
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```


# Discussion

## Total Outbreak, Coronavirus Outbreak, and Respiratory Outbreak {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Influenza Outbreak & Other 

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Other outbreaks


## Weaknesses and next steps

One main weakness with this analysis is it is limited to healthcare facilities, which fail to be representative of the larger demographic of Toronto. A next step would be to analyze disease outbreak in the general population.

\newpage


\newpage


# References


